---
title: "SCI6245: Clinical Bioinformatics"
subtitle: "*Module 1: Reproducible, Ethical and Effective Clinical Bioinformatics*"
author: "Anais Chen <br> Master of Bioinformatics Student <br> Edith Cown University <br> achen3@our.ecu.edu.au"
date: "Created: 25th February 2021 <br> Last Updated: `r format(Sys.Date(), format = '%d %B %Y')`"
output: 
  html_document:
    css: buttons.css
    includes:
      before_body: header.html
    toc: TRUE
    toc_depth: 2
    toc_float: 
      collapsed: false
      smooth_scroll: true
---
<script src="hideOutput.js"></script>   
## Overview
Module 1 discusses conduct in clinical bioinformatics research practices, including reproducibility,
ethics, and effective communication. 

```{r setup, include = FALSE }
knitr::opts_chunk$set(echo = TRUE)
```
## Load Dependendies
```{r, echo=FALSE}
library(tidyverse)
library(readxl)
library(kableExtra)
library(ggpubr)
```

## Data Preparation
```{r}
#lists info sheets
excel_sheets("./data/surveys_data.xlsx")

#loading the main data
survey_data <- read_excel("./data/surveys_data.xlsx", sheet = 1)

#loading the plot info
plot_info <- read_excel("./data/surveys_data.xlsx", sheet = 2)

#load species info
species_info <- read_excel("./data/surveys_data.xlsx", sheet = 3)
```
First, I clean up the sex data and give new, easier to understand labels.
<div class="fold o">
```{r}
#checks data structure, shows variable names and some code
str(survey_data)

#recodes sex to numeric
survey_data$sex <- ifelse(survey_data$sex=="M", 0, ifelse(survey_data$sex == "F", 1, "NA"))
survey_data$sex <- factor(survey_data$sex, 
                          levels=c(0,1), labels=c("Male", "Female"))
```
Then I label the plots by the plot_type information.

```{r}
#explores the data structure
str(plot_info)

#label plot_id by the plot_type info
survey_data$plot_id <- factor(survey_data$plot_id, 
                              levels=plot_info$plot_id, 
                              labels=plot_info$plot_type)
```
<br>
To complete the data labels, I create a new variable for species ID, combining the genus and the species variables, and append this to the survey data. 
<br>
```{r}
#explores data structure
str(species_info)

#merges genus with species with a space between them
species_info$species_lab <- paste(species_info$genus, species_info$species, " ")

#checks first 10 rows
head(species_info[1:10,])

#removes lagging space
species_info$species_lab <- trimws(species_info$species_lab, which = "right")

#check first 10 rows
head(species_info[1:10,])

#labels species_id values in survey_data
survey_data$species_id <- factor(survey_data$species_id, 
                                 levels=species_info$species_id,
                                 labels=species_info$species_lab)

#checks main data
str(survey_data)

#adds genus and taxa to survey_data
survey_data <- merge(survey_data, species_info[,c(2,4,5)], 
                     by.x="species_id", by.y="species_lab")

#checks main data
str(survey_data)
```
</div>
<br>
On inspection of the data, many data points seem to be missing weight, hindfoot length and sex observations, so I flag the data with missing inputs and create a new survey data set that does not include them. I then check to make sure all the missing data points have been excluded.
<br>
```{r}
#creates new data set without missing data for sex or weight
survey_data$sexNA <- ifelse(is.na(survey_data$sex), 1, 0)  #creates number index for missing sex
survey_data$weightNA <- ifelse(is.na(survey_data$weight), 1, 0)
survey_data$anyNA <- ifelse(survey_data$sexNA==1 | survey_data$weightNA==1, 1, 0)  #creates number index for either missing

table(survey_data$sexNA, survey_data$anyNA)    #checks all sexNA are counted
```
```{r}
table(survey_data$weightNA, survey_data$anyNA) #checks all weightNA are counted
```

```{r}
table(survey_data$anyNA)
```
```{r}
survey_data2 <- survey_data[!survey_data$anyNA==1, ]    #creates new data excluding missing
dim(survey_data)

```
<div class="fold o">
```{r}
str(survey_data2)
```

```{r}
head(survey_data2)
```
</div>
<br>
Finally, I create a new variable to separate the data into bins by year, with a 5 year interval, for exploratory analysis.
<br>
```{r}
#creates a variable for year bins in intervals of 5
yearBreaks <- c(min(survey_data2$year), 1982, 1988, 1992, 1997, max(survey_data2$year))

#buckets the values into bins
survey_data2$yearBin <- cut(survey_data2$year,
                breaks = yearBreaks,
                include.lowest = TRUE,
                right = FALSE,
                labels = c("1977-1982", "1982-1987", "1987-1992", "1992-1997", "1997-2002"))

#checks year bin variable

table(survey_data2$yearBin)
```

## Exploratory Analysis {.tabset .tabset-fade}

### Preparation
<div class="fold s">
```{r}
#creates total count variables
yearSumm <- survey_data2 %>%
  group_by(yearBin) %>%
  count() %>%
  as.data.frame()
yearSumm$p <- round(yearSumm$n/nrow(survey_data2)*100, 1) #calculates percentages

#creates sex count variables
sexSumm <- survey_data2 %>%
  select(yearBin, sex) %>%
  group_by(yearBin) %>%
  count(sex) %>%
  as.data.frame()
sexSumm$p <- c(round(sexSumm$n[1:2]/sum(sexSumm$n[1:2])*100, 1),
               round(sexSumm$n[3:4]/sum(sexSumm$n[3:4])*100, 1),
               round(sexSumm$n[5:6]/sum(sexSumm$n[5:6])*100, 1),
               round(sexSumm$n[7:8]/sum(sexSumm$n[7:8])*100, 1),
               round(sexSumm$n[9:10]/sum(sexSumm$n[9:10])*100, 1)) #calculates percentages

#creates mean variables
meanSumm <- survey_data2 %>%
  select(yearBin, hindfoot_length, weight) %>%
  group_by(yearBin) %>%
  summarise(across(.cols = everything(), ~mean(.,na.rm = TRUE))) %>%
  as.data.frame()

#creates sd variables
sdSumm <- survey_data2 %>%
  select(yearBin, hindfoot_length, weight) %>%
  group_by(yearBin) %>%
  summarise(across(.cols = everything(), ~sd(.,na.rm = TRUE))) %>%
  as.data.frame()

#numbers in plots by yearBin
plotSumm <- survey_data2 %>%
  select(yearBin, plot_id) %>%
  group_by(yearBin) %>%
  count(plot_id) %>%
  as.data.frame()
 #calculates plot percentages
plotSumm$p <- c(round(plotSumm$n[1:5]/sum(plotSumm$n[1:5])*100, 1),
                round(plotSumm$n[6:10]/sum(plotSumm$n[6:10])*100, 1),
                round(plotSumm$n[11:15]/sum(plotSumm$n[11:15])*100, 1),
                round(plotSumm$n[16:20]/sum(plotSumm$n[16:20])*100, 1),
                round(plotSumm$n[21:25]/sum(plotSumm$n[21:25])*100, 1))
```
```{r}
#creates a dataframe with the desored formatting of the final table
t1 <- data.frame(vars = c(paste0("**Total Count** (n) ","*(%)*"),
                          paste0("**Sex** (n) ","*(%)*"),
                          "*Male", "Female",
                          paste0("**Hindfoot Length** (mm) ","mean *(SD)*"),
                          paste0("**Weight** (g) ","mean *(SD)"),
                          paste0("**Plot Information** (n) ","*(%)*"),
                          "*Spectab Exclosure*", "*Control*", "*Rodent Exclosure*",
                          "*Long-term Krat Exclosure*", "*Short-term Krat Exclosure*"),
#column for 1977-1982
                 bin1 = c(paste0(yearSumm$n[1], " (", yearSumm$p[1], "%)"), "",
                          paste0(sexSumm$n[1:2], " (", sexSumm$p[1:2], "%)"),
                          paste0(round(meanSumm$hindfoot_length[1], 2), " (",
                                 round(sdSumm$hindfoot_length[1], 2), ")"),
                          paste0(round(meanSumm$weight[1], 2), " (",
                                 round(sdSumm$weight[1], 2), ")"), "",
                          paste0(plotSumm$n[1:5], " (", plotSumm$p[1:5], ")")),
#column for 1982-1987
                 bin2 = c(paste0(yearSumm$n[2], " (", yearSumm$p[2], "%)"), "",
                          paste0(sexSumm$n[3:4], " (", sexSumm$p[3:4], "%)"),
                          paste0(round(meanSumm$hindfoot_length[2], 2), " (",
                                 round(sdSumm$hindfoot_length[2], 2), ")"),
                          paste0(round(meanSumm$weight[2], 2), " (",
                                 round(sdSumm$weight[2], 2), ")"), "",
                          paste0(plotSumm$n[6:10], " (", plotSumm$p[6:10], ")")),
#column for 1987-1992
                 bin3 = c(paste0(yearSumm$n[3], " (", yearSumm$p[3], "%)"), "",
                          paste0(sexSumm$n[5:6], " (", sexSumm$p[5:6], "%)"),
                          paste0(round(meanSumm$hindfoot_length[3], 2), " (",
                                 round(sdSumm$hindfoot_length[3], 2), ")"),
                          paste0(round(meanSumm$weight[3], 2), " (",
                                 round(sdSumm$weight[3], 2), ")"), "",
                          paste0(plotSumm$n[11:15], " (", plotSumm$p[11:15], ")")),

#column for 1992-1997
                 bin4 = c(paste0(yearSumm$n[4], " (", yearSumm$p[4], "%)"), "",
                          paste0(sexSumm$n[7:8], " (", sexSumm$p[7:8], "%)"),
                          paste0(round(meanSumm$hindfoot_length[4], 2), " (",
                                 round(sdSumm$hindfoot_length[4], 2), ")"),
                          paste0(round(meanSumm$weight[4], 2), " (",
                                 round(sdSumm$weight[4], 2), ")"), "",
                          paste0(plotSumm$n[16:20], " (", plotSumm$p[16:20], ")")),
#column for 1992-1997
                 bin5 = c(paste0(yearSumm$n[5], " (", yearSumm$p[5], "%)"), "",
                          paste0(sexSumm$n[9:10], " (", sexSumm$p[9:10], "%)"),
                          paste0(round(meanSumm$hindfoot_length[5], 2), " (",
                                 round(sdSumm$hindfoot_length[5], 2), ")"),
                          paste0(round(meanSumm$weight[5], 2), " (",
                                 round(sdSumm$weight[5], 2), ")"), "",
                          paste0(plotSumm$n[21:25], " (", plotSumm$p[21:25], ")")),
stringsAsFactors = FALSE)
```
</div>
This checks the structure of the data frame before formatting into a table with kable.
<div class="fold o">
```{r}
t1
```
</div>
<br>
Then, I create summary plots of the data variables
<br>
<div class="fold s">
```{r}
#removes any levels where species_id has no data points
survey_data2$species_id <- droplevels(survey_data2$species_id) 
survey_data2$genus <- as.factor(survey_data2$genus)

#creates plot for samples collected each year by genus
p1 <- survey_data2 %>%
  select(year, genus) %>%
  group_by(year) %>%
  count(genus) %>%
  ggplot( aes(x=year, y=n, colour=genus)) +
  geom_line(size=1.2) +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  xlab("Year") + ylab("Number of Samples Collected") 

#creates a boxplot comparing the weight of the sex variable by plot_id
p2 <- ggplot(survey_data2, aes(x=as.factor(plot_id), y=weight, fill = as.factor(sex))) +
    geom_boxplot(size = 0.5,
                 alpha = 0.5) +
  theme_bw() +
  labs(y = "Weight (g)", x = "", fill = "Sex") + theme(legend.position = "none") +
coord_flip()

#creates a boxplot comparing the hindfoot length of the sex variable by plot_id
p3 <- ggplot(survey_data2, aes(x=as.factor(plot_id), y=hindfoot_length, fill = as.factor(sex))) +
    geom_boxplot(size = 0.5,
                 alpha = 0.5) +
  theme_bw() +
  labs(y = "Hindfoot Length (mm)", x = "", fill = "Sex") +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
coord_flip()

#creates plots for weight and length by genus for each plot
p4 <- ggplot(survey_data2, aes(x=hindfoot_length, y=weight, colour=as.factor(genus))) +
  geom_point(na.rm = TRUE, alpha=0.5) + theme_bw() +
  xlab("Hindfoot Length (mm)") + ylab("Weight (g)") + 
  facet_grid(~ plot_id) + 
  scale_colour_discrete(name = "Genus") + theme(legend.position = "none")
  
#creates plots of weight for each plot
p5 <- ggplot(survey_data2, aes(x = weight, fill = plot_id)) +
  geom_histogram( aes(y = ..density.. ), bins = 50, colour = "black", alpha = 0.5) +
  geom_density(size=1, alpha=0.4) + 
  theme_bw() + facet_grid(~ plot_id) + theme(legend.position = "none") +
  xlab("Weight (g)") + ylab("Density")

#creates plots of hindfoot length for each plot
p6 <- ggplot(survey_data2, aes(x = hindfoot_length, fill = plot_id)) +
  geom_histogram( aes(y = ..density.. ), bins = 50, colour = "black", alpha = 0.5) +
  geom_density(size=1, alpha=0.4) + 
  theme_bw() + facet_grid(~ plot_id) + theme(legend.position = "none") +
  xlab("Hindfoot Length (mm)") + ylab("Density")
  
```
</div>
### Summary Table
<div class="fold s">
```{r}
#creates summary table with kable
kable(t1, 
      caption = "**Table 1. Summary of study variables in 5 year intervals**", #table caption
      col.names = c("", "1977-1982", "1982-1987", "1987-1992", "1992-1997", "1997-2002"), #column titles, leaving blank space for first column
      align = "lccccc",
      type = "") %>%
  column_spec(1, width_min = "4cm", border_right = "2px solid black") %>%
  column_spec(c(2,3,4,5,6), width_min = "3cm", border_right = "1px solid #dddddd" ) %>%
  row_spec(0, bold=T, color="white", background="#666666") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  add_indent(c(3,4,8:12)) %>%  # indent rows for measures with levels
  footnote(general = c("*% are within groups for all non-missing values*")) %>% # add footnote
  add_header_above(c(" " = 1, "Year" = 5), bold=T, color="white", background="#666666", include_empty = T)
```
</div>
### Data Visualisation 
```{r}
ggarrange(
  p4, p1,
  labels = c("a", "b"),
  nrow = 2, heights = c(1,1.5)) #copied from Alyce's example
```
<center><font size = "3">**Figure 1. Visualisations of Genera. a)** Hindfoot Length vs Weight, **b)** Samples Collected</font></center>
```{r}
ggarrange( 
  p2, p3,
  labels = c("a", "b"),
  ncol = 2, nrow = 1)
```
<center><font size = "3">**Figure 2. Distributions of Sex. a)** By Weight, **b)** By Hindfoot Length</font></center>
```{r}
ggarrange(
  p5, p6,
  labels = c("a", "b"),
  ncol = 1, nrow = 2
)
```
<center><font size = "3"> **Figure 3. Distribution of Exclosure. a)** By Weight, **b)** By Hindfoot Length</font></center>


## Session Information
***
```{r}
sessionInfo()
```